Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-11 paper/sections/metis-language-and-proofs.tex
[ paper ] perliminaries added.

\name{Isabelle/HOL} incorporated \Metis as a macro-step
reconstruction tool for explaining proof steps (usually \CNF goals)
replied from other \ATPs like \name{CVC4}, \name{Vampire}, or
\name{Z3} (more details in Ref.~\cite{paulson2007source}).
In \cite{Farber2015}, \Metis was used to provide a tactic for
\name{HOL Light}; this challenges others like \name{MESON}
or the \name{leanCoP} tactic that reconstructs proofs
delivered by \name{leanCoP}~\cite{Farber2016}.

@@

The \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula
assumes one \name{role}, this could be an \name{axiom},
\name{conjecture}, \name{definition}, or a \name{hypothesis}.
The formulas include Boolean connectives (\verb!&!, \verb!|!,
\verb!=>!, \verb!{<=>!, \verb!~!) and logic constants \verb!$true!
and \verb!$false!.

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (see. e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.\ref{fig:metis-example}).

@@

This lack of the \Metis' documentation is an important matter for
us in the proof-reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior has become a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

>>>>

\name{Isabelle/HOL} incorporated \Metis as a macro-step
reconstruction tool for explaining proof steps (usually \CNF goals)
replied from other \ATPs like \name{CVC4}, \name{Vampire}, or
\name{Z3} (more details in Ref.~\cite{paulson2007source}).
In \cite{Farber2015}, \Metis was used to provide a tactic for
\name{HOL Light}; this challenges others like \emph{MESON}
or the \emph{leanCoP} tactic that reconstructs proofs
delivered by \name{leanCoP}~\cite{Farber2016}.

@@

The \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula
assumes one \name{role}, this could be an \name{axiom},
\name{conjecture}, \name{definition}, or a \name{hypothesis}.
The formulas include Boolean connectives (\verb!&!, \verb!|!,
\verb!=>!, \verb!<=>!, \verb!~!) and logic constants \verb!$true!
and \verb!$false!, for the symbols (∧, ∨, ⇒, ⇔, ¬, ⊤, ⊥), respectively.

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (see. e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.~\ref{fig:metis-example}).

@@

This lack of the \Metis' documentation is an important matter for
us in the proof-reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior is a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

### 2017-09-11 paper/sections/proof-reconstruction.tex
added text.
[ paper ] perliminaries added.

\section{Proof Reconstruction}
\label{sec:proof-reconstruction}

@@

% node[below] {send to}
\draw[->, thick] (problem) to
  % node[below] {\tiny encoding}
  (tptp);
\draw[->, thick] (tptp) to
  % node[below] {\tiny }
  (metis);
\draw[->, thick] (metis) to
  % node[below] {\tiny replies on}
  (tstp);
\draw[->, thick] (tstp) to
  % node[right] {\tiny parsing}
  (athena);
\draw[->, thick] (athena) to
  % node[below] {\tiny traslation}
  (agdafile);
\draw[->, thick] (agdafile) to
  % node[below] {\tiny type-checking}
  (agda);
\draw[->, thick] (agda) to (verified);
\draw[->, thick, gray] (agda) to (failure);
\end{tikzpicture}
\caption{Proof reconstruction overview.}
\label{fig:proof-reconstruction-workflow}
\end{figure}

The proof reconstruction approach proposed here consists of a series 
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source 
system to a target system. In our case, the system of origin, the 
automatic theorem prover, is \Metis; the target system is the proof 
assistant, \Agda. We choose \Agda, but another proof assistant with 
support of type theory and features similar like \Agda is valid; our 
results do not depend on the proof assistant but the support of inductive data types, and termination checking, and type-checking.
\improvement{Make this statement stronger and clear.}

@@

Finally, we use the proof assistant in step No. 7 to type-check the 
proof and verify its correctness. We have only one answer from two 
possible answers: type-checking succeed (i.e., the proof is 
valid) or type-checking failed.
If the type-check fails, one of the three actors are 
responsible: \Metis, \Athena, or \Agda. We are responsible for 
errors committed in the translation from \TSTP to \Agda; lack of the 
coverage of the \Metis' inference rules, or bad assumptions about 
them.

@@

>>>>

\section{Proof-Reconstruction}
\label{sec:proof-reconstruction}

\subsection{Workflow}
\label{ssec:workflow}

@@

% node[below] {send to}
\draw[->, thick] (problem) to
  % node[below] {\tiny encoding}
  (tptp);
\draw[->, thick] (tptp) to
  % node[below] {\tiny }
  (metis);
\draw[->, thick] (metis) to
  % node[below] {\tiny replies on}
  (tstp);
\draw[->, thick] (tstp) to
  % node[right] {\tiny parsing}
  (athena);
\draw[->, thick] (athena) to
  % node[below] {\tiny traslation}
  (agdafile);
\draw[->, thick] (agdafile) to
  % node[below] {\tiny type-checking}
  (agda);
\draw[->, thick] (agda) to (verified);
\draw[->, thick, gray] (agda) to (failure);
\end{tikzpicture}
\caption{proof-reconstruction overview.}
\label{fig:proof-reconstruction-workflow}
\end{figure}

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but another proof assistant with
support of type theory and features similar like \Agda is valid; our
results do not depend on the proof assistant but the support of
inductive data types, termination checking, and type-checking.

@@

Finally, we use the proof assistant in step No. 7 to type-check the
proof and verify its correctness. Next, we only have one answer:
type-checking succeed (i.e., the proof is valid) or type-checking failed.
If the type-check fails, one of the three actors in the process are
responsible: \Metis, \Athena, or \Agda. We are responsible for
errors committed in the translation from \TSTP to \Agda; lack of the
coverage of the \Metis' inference rules, or bad assumptions about
them. \improvement{Make this statement stronger and clear.}

@@

% ...................................................................

% \footnote{We use the \Agda standard library's implementation of the List data type.}.

\subsection{A Brief Review of Agda}
\label{ssec:agda-definition}

\Agda is an interactive system for constructing proofs and programs,
based on Martin-L\"{o}f's type theory and extended with records, parametrised
modules, among other features.
One of the main strengths of \Agda is its support for writing proofs,
which we shall call \Agda's proof engine. It consists of: support for
inductively defined types, including inductive families, and function
definitions using pattern matching on such types, normalisation during type-
checking, commands for refining proof terms, coverage checker and termination
checker.

The inductive approach for representing classical propositional logic
is better because we benefit from \Agda's proof engine and its Unicode
support that allows us writing proofs similar as we find in math text books.

In \Agda, we define a proposition formula
as an inductive type using the keyword \texttt{data}.
The \name{PropFormula} data type stands for propositions,\cite{Altenkirch2015}.
We include every connective, logic constant as a constructor.
We also define the constructor for atomic propositions \verb!Var!
to represent based on a natural number $i$, the atomic proposition
$p_i$ for a finite universe of them ($i \leq n$), this is $Fin n$ in the
definition.

\begin{verbatim}
data PropFormula : Set where
  Var : Fin n → PropFormula
  ⊤ : PropFormula
  ⊥ : PropFormula
  _∧_ _∨_ _⇒_ _⇔_ : (φ ψ : PropFormula) → PropFormula
  ¬_ : (φ : PropFormula) → PropFormula
\end{verbatim}

\begin{remark}
In the above definition, we defined all connectives, execept negation, as
a binary connective. In \Agda, we named constructors for those functions
that inhabit the data type, for example, the boolean connecetives and
the function for atomic propositions, \verb!Var!, are constructors.
The underscores in a definition like
\begin{verbatim}
_⇔_ : (φ ψ : PropFormula) → PropFormula
\end{verbatim}
stands for an argument, and \verb!(φ ψ : PropFormula)! intend for the
uncurryfied version of the type:
$$PropFormula → PropFormula → PropFormula$$
\end{remark}

\begin{definition}(Negative Normal Form (NNF))
Una forma normal negativa para la formula $\varphi$ es aquella
en la cual negaciones aparecen solamente en los literales y
la expresión se encuentra en términos unicamente
de los operadores booleanos ($\neg$, $\wedge$, $\vee$).
\end{definition}

Para transformar una formula $\varphi$ a su forma equivalente NNF,
aplicamos recursivamente las siguientes ecuaciones que definen
la función \name{nnf}.

\begin{equation}
\label{eq:nnf-definition}
\nnf(φ) =
\begin{cases}
\nnf(φ₁) ∧ \nnf(φ₂)
   & φ := φ₁ ∧ φ₂  \\

\nnf(φ₁) ∨ \nnf(φ₂)
   & φ := φ₁ ∨ φ₂  \\

\nnf(¬ φ₁ ∨ φ₂)
   & φ := φ₁ ⇒ φ₂  \\

\nnf(φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁)
   & φ := φ₁ ⇔ φ₂\\

\nnf(¬ φ₁ ∨ ¬ φ₂)
   & φ := ¬ (φ₁ ∧ φ₂) \\

\nnf(¬ φ₁ ∧ ¬ φ₂)
   & φ := ¬ (φ₁ ∨ φ₂) \\

\nnf(φ₁)
   & φ := ¬ ¬ φ₁     \\

\nnf(¬ (φ₂ ∨ ¬ φ₁))
   & φ := ¬ (φ₁ ⇒ φ₂)  \\

\nnf(¬ (φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁))
   & φ := ¬ (φ₁ ⇔ φ₂) \\

⊥  & φ := ¬ ⊤        \\

⊤  & φ := ¬ ⊥        \\

φ  & \text{otherwise}
\end{cases}
\end{equation}

\begin{definition}
Una funcion recursiva se dice bien estructura, si el tamaño de los argumentos
con que con esta se define se reducen en tamaño por cada llamada a la
funcion.
\end{definition}

De la anterior definincion, sintacticamente la funcion nnf no es recursivamente
bien estructurado. El problem que no lo sea esta en nuestra traducion
de este tipo de funciones a la teoria de tipos.

Veamos otros ejemplos de funciones con el mismo problema que utilizaremos mas
adelante para formalizar la inferencia strip.

\begin{equation}
\label{eq:unshunt-definition}
\unshunt(φ) =
\begin{cases}
\unshunt((φ₁ ∧ φ₂) ⇒ φ₃)
  & φ := φ₁ ⇒ (φ₂ ⇒ φ₃)\\

\unshunt(φ₁ ⇒ φ₂) ∧ \unshunt(φ₁ ⇒ φ₃)
  & φ := φ₁ ⇒ (φ₂ ∧ φ₃)\\

φ &\text{otherwise}
\end{cases}
\end{equation}

Para traducir las funciones en un manera que esten bien definidas,
utilizamos el metodo propuesto por Ana Bove, el cual basicamente consiste

una funcion general recursiva $f:\sigma \to \tau$, su formalizacion en
teoria de tipos consiste en definir un predicado inductivo $Ind$,
$P:\sigma \to Ind$

En nuestro caso, el conjunto de los numeros servira como nuestro
conjunto inductivo y como segundo argumento para nueva funcion.
De manera, que la funcion g estara definida por recursion estructural en
su segundo termino.

Para aplicar la tecnica antes mencionada para nuestras funciones,
veamos su aplicación con la funcion \unshunt porque para las demas,
seguiremos un camino similar, f con tipo:

\[
\unshunt^{*} : PROP \to \mathcal{N} \to PROP
\]

\begin{equation}
\label{eq:unshunt-structured}
\unshunt^{*}(φ, n) =
\begin{cases}
\unshunt^{*}((φ₁ ∧ φ₂) ⇒ φ₃, n-1)
  & φ := φ₁ ⇒ (φ₂ ⇒ φ₃)\\

\unshunt^{*}(φ₁ ⇒ φ₂, n-1) ∧ \unshunt^{*}(φ₁ ⇒ φ₃,n-1)
  & φ := φ₁ ⇒ (φ₂ ∧ φ₃)\\

φ & n := 0
\end{cases}
\end{equation}

\begin{definition}
El rango $rank(\phi)$ de una proposition $\phi$ es definido como

\begin{equation}
\label{eq:rank-definition}
rank(\phi)=
\begin{cases}
0 &\text{for atomic }\phi\\

\max(rank(\phi_1),rank(\phi_2)) + 1
  & \phi := \phi_1\,\square\, \phi_2 \\
rank(\phi_1) + 1  & \phi := \neg \phi_1
\end{cases}
\end{equation}

\end{definition}

Then, we our final revision for the function $\unshunt$ is the
function$\unshunt^{+}$ with type $PROP \to PROP$
with a structured recursion,

\begin{equation}
\unshunt^{+}(\varphi) = \unshunt^{*}(\varphi, rank(\varphi))
\end{equation}

In the following section, our task will consist of a searching
appropriate functions like $rank$ to assure the termination of
our transformation functions like $nnf$.

% ................................................

\subsection{Splitting the goal}

\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂)
  & φ := φ₁ ∧ φ₂\\

\unshunt(¬ φ₁ ⇒ \mysplit(φ₂))
  & φ := φ₁ ∨ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(φ₂))
  & φ := φ₁ ⇒ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(φ₂)) ∧ \unshunt(φ₂ ⇒ \mysplit(φ₁))
  & φ := φ₁ ⇔ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ∧ φ₂)\\

\unshunt(\mysplit(¬ φ₁)) ∧ \unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ∨ φ₂)\\

\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ⇒ φ₂)\\

\unshunt(φ₁ ⇒ \mysplit(¬ φ₂)) ∧ \unshunt(¬ φ₂ ⇒ \mysplit(φ₁))
  & φ := ¬ (φ₁ ⇔ φ₂)\\

\unshunt(\mysplit(φ₁)
  & φ := ¬ ¬ φ₁\\

⊤ & φ := ¬ ⊥\\

⊥ & φ := ¬ ⊤\\

φ &\text{otherwise}
\end{cases}
\end{equation}
